Interestingness Measures for Theory Exploration

### 2015-11-23 TheoryExploration/interesting.tex
Keep interesting.tex in git

In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain. To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}. Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.

@@

We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context. We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape. A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}. We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness. Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.

We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.

\begin{description}
\item{Interestingness}
  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}. Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
\end{description}

@@

\section{Exploration in Theorem Proving}
\label{examples}

@@

Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures. See \S \ref{auxiliarylemmas} for example definitions of each type.}

@@

\subsection{Auxiliary Lemmas} \label{auxiliarylemmas}

@@

\subsection{Tests} \label{tests}

@@

\section{Related Work}
\label{related}

@@

As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do. The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible. The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.

@@

\paragraph{Feature Extraction}\label{featureextraction}

>>>>

In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain. To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{sec:haskell}. Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.

@@

We begin our undertaking in \S \ref{sec:background} by introducing the Haskell language, as well as the relevant fields of verification for context. We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape. A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{sec:related}. We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness. Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.

We discuss our present contributions in \S \ref{sec:current} and future research directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.

\begin{description}
\item{Interestingness}
  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
\end{description}

@@

\section{Exploration in Theorem Proving}
\label{sec:examples}

@@

Generalisation also occurs frequently when reasoning about \emph{tail-recursive} definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can be executed in constant space using a loop, whereas recursion in non-tail positions may require a growing number of stack frames or nested closures. See \S \ref{sec:auxiliarylemmas} for example definitions of each type.}

@@

\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}

@@

\subsection{Tests} \label{sec:tests}

@@

\section{Related Work}
\label{sec:related}

@@

As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do. The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible. The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \cite{berlyne1970novelty}.

@@

\paragraph{Feature Extraction}\label{sec:featureextraction}
